---
title: "Exploring the BRFSS data"
author: "Hartmut Schaefer"
date: 2024-01-13
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
    code_folding: show
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```
\

## Setup

### Load packages

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
library(ggmosaic)
library(RColorBrewer)
```

### Load data

Load data set

```{r load-data}
load("~/R/StatRWeb/data/brfss2013.RData")
```
\


### Prepare data set

Rename data set and convert to tibble for easier handling

```{r}
df_raw <- brfss2013
```

Add a row ID to each observation

```{r}
df_raw <- df_raw %>% 
  rowid_to_column(var="id_raw")
```


* * *

## Part 1: Data

### Data exploration

The objective of BRFSS data collection is to determine behavioral risk factors to chronic diseases in adults in the US.

The data were collected through telephone interviews in 51 states of the US in year 2013 by stratified sampling, i.e. random samples for each strata (state).
There are 491,775 observations in the data set. The sample is large enough to generalize to the population.

Limitations: \
The survey does not include institutionalized adults. This may cause some bias and affect the conclusion.

The survey uses a random sample and therefore only correlations can be inferred between variables. In order to infer causality a controlled experiment would be required.



* * *

## Part 2: Research questions

**Research question 1:**

Is income level (`income2`) correlated with general health (`genhth`)?\

a. What is the probability of good health given a high income?\
b. What is the probability of high income given a good health?\
c. Are the variables income level and general health independent or dependent?\


**Research question 2:**

Is general health (`genhlth`) correlated with physical activity (`pa1min_`) and how does it change by gender (`sex`)?\

a. Describe physical activity statistically\
b. Find the relationship between general health and physical activity\
c. Is there a difference in this relationship (b) in respect to gender?\


**Research question 3:**

Is physical activity (`pa1min_`) correlated with BMI (Body Mass Index) (`X_bmi5`) and how does it change with general health (`genhlth`) and gender (`sex`)?\

a. Describe BMI statistically\
b. Find the relationship between physical activity and BMI\
c. Is there a difference in this relationship (b) in respect to general health and gender?



* * *

\

### Research question 1:

Is the income level (`income2`) correlated with general health (`genhlth`)?\

a. What is the probability of good health given a high income?\
b. What is the probability of high income given a good health?\
c. Are the variables income level and general health independent or dependent?\

Here we are comparing two categorical variables. For the comparison we will create a **contingency table** and visualize the proportions between each category.\
Next we will calculate **conditional probabilities** and test whether both categories are independent.\
\
\

#### Preparing the data set and data cleaning

**Reduce the number of variables**: Extract variable `genhth` (general health) and `income2` (income level)

```{r data cleaning q1}
df_select_q1 <- df_raw %>% 
  select(income2, genhlth)
```
\

**Inspecting the categories** for variable `genhlth`:

```{r NA check genhlth}
# identify NAs in genhlth
df_select_q1 %>% 
  group_by(genhlth) %>% 
  summarise(count=n())
```
There are five ordinal categories: excellent, very good, good, fair, poor and a number of NAs
\


**Inspect the categories** for variable `income2`:

```{r NA check income2}
# identify NAs in income2
df_select_q1 %>% 
  group_by(income2) %>% 
  summarise(count=n())
```
There are 7 ordinal categories: from <$5K to >$75K and a number of NAs.
\
\

**Remove NAs** form the data set

Rows will be deleted if any of the variable contains a NA (not available). We are using `!is.na()` function, i.e. "is not NA". Note: logical expression "A or B" is equivalent with "NOT A and NOT B".

```{r remove NAs q1}
# remove NAs
df_q1 <- df_select_q1 %>% 
  group_by(genhlth,income2) %>% 
  filter(!is.na(genhlth), !is.na(income2))
```


**Rename values** in category `income2` to simplify output. The levels (names) in the categorical variable `income2` are shorten for easier plotting, using function `recode()`

```{r rename values in income2}
# Rename values in income2
df_q1$income2 <- recode(df_q1$income2, 
                          "Less than $10,000" = "~$10K",
                          "Less than $15,000" = "~$15K",
                          "Less than $20,000" = "~$20K",
                          "Less than $25,000" = "~$25K",
                          "Less than $35,000" = "~$35K",
                          "Less than $50,000" = "~$50K",
                          "Less than $75,000" = "~$75K",
                          "$75,000 or more" = "$75K~")
```
\
\

#### Exploring the data set

**Visualize distribution** of categories using bar-plots

General Health:

```{r distribution cat values genhlth}
# Bar chart of category genhelth (COUNT)
df_q1 %>% 
  ggplot(aes(x = genhlth)) +
  geom_bar(fill="gray", color="white")+
  labs(title = "General health: Distribution of counts")
```
\
Highest counts are in categories "very good" and "good"\
\

Income Level:

```{r distribution cat values income levels}
# Bar chart of category income levels (COUNT)
df_q1 %>% 
  ggplot(aes(x = income2)) +
  geom_bar(fill="gray", color="white")+
  labs(title = "Income levels: Distribution of counts")
```
\
Highest count in category "$75K or more". Indication of certain bias to more respondents in higher income level. The relationship between "income" and "general health" maybe biased as well. Nevertheless, we will proceed with the investigation.
\
\

**Create a contingency table**

```{r creating contingency table}
# Creating a contingency table with absolute counts
tb_cont_table1_q1 <- table(df_q1)
tb_cont_table1_q1
```
\


**Visualize the contingency table** as mosaic plot:

```{r make mosaic plot, fig.width=8, fig.asp=0.6, warning=FALSE}

# change format from table to data frame (tibble)
df_cont_table1_q1 <- as_tibble(tb_cont_table1_q1)

# change character format to factor
df_cont_table1_q1$genhlth <- factor((df_cont_table1_q1$genhlth),
                                    levels = c("Poor",
                                               "Fair",
                                               "Good",
                                               "Very good",
                                               "Excellent"))

df_cont_table1_q1$income2 <- factor((df_cont_table1_q1$income2),
                                    levels = c("~$10K",
                                               "~$15K",
                                               "~$20K",
                                               "~$25K",
                                               "~$35K",
                                               "~$50K",
                                               "~$75K",
                                               "$75K~"))
                                      
# creating a mosaic plot
df_cont_table1_q1 %>% 
  ggplot()+
  geom_mosaic(aes(x = product(income2,genhlth), fill=income2, weight = n))+
  coord_flip()+
  labs(title = "Relations between income level and general health status",
       fill = "Income levels")+
  ylab("Income Level (proportions)")+
  xlab("General Health (proportions)")+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())+
  scale_fill_brewer(palette = "RdYlBu")
```
\

Conclusion: \

- the higher the "income" level the better the  the general health status\
- the lower the "income" level the poorer the health status\
- in the mid-income range the health status is more equally distributed\

From this graph it is relatively difficult to grasp, that the income level is related to health status. We can simplify the graph by aggregating the levels to two categories per variable\
\


**Simplify the categories** by aggregation

For better extraction of the main trend, the income levels will be aggregated into two groups: income "above median" and "below median". As threshold we will use the national (US) median income in 2013. According to the census bureau the median income in 2013 was $52,250. Therefore we will cut into two groups: "<$50K" and ">$50K".

The general health status will be also aggregated into two groups: Health "Less good" and "More than good".

**Aggregate** levels:

```{r aggregate levels}
df_cont_table1_agg_q1 <- df_q1


df_cont_table1_agg_q1$income2 <- recode(df_cont_table1_agg_q1$income2, 
                        "~$10K" = "below median",
                        "~$15K" = "below median",
                        "~$20K" = "below median",
                        "~$25K" = "below median",
                        "~$35K" = "below median",
                        "~$50K" = "below median",
                        "~$75K" = "above median",
                        "$75K~" = "above median")

df_cont_table1_agg_q1$genhlth <- recode(df_cont_table1_agg_q1$genhlth,
                          "Poor" = "less good",
                          "Fair" = "less good",
                          "Good" = "less good",
                          "Very good" = "more than good",
                          "Excellent" = "more than good")
```
\
\

**Creating the contingency table from aggregates**

```{r creating contingency table 2}
# Creating a contingency table with absolute counts
tb_cont_table1_agg_q1 <- table(df_cont_table1_agg_q1)

tb_cont_table2_agg_q1 <- addmargins(tb_cont_table1_agg_q1)
tb_cont_table1_agg_q1
```
\

**Visualize the contingency table** as mosaic plot:

```{r make mosaic plot 2, fig.width=8, fig.asp=0.5}

# change from table to data frame
df_cont_table1_agg_q1 <- as_tibble(tb_cont_table1_agg_q1)

# change character to factor
df_cont_table1_agg_q1$genhlth <- factor((df_cont_table1_agg_q1$genhlth),
                                    levels = c("less good",
                                               "more than good"))

df_cont_table1_agg_q1$income2 <- factor((df_cont_table1_agg_q1$income2),
                                    levels = c("below median",
                                               "above median"))
                                      
# creating a mosaic plot
df_cont_table1_agg_q1 %>% 
  ggplot()+
  geom_mosaic(aes(x = product(income2,genhlth), fill=income2, weight = n), alpha=0.5)+
  coord_flip()+
  labs(title = "Relations between income level and general health status",
       subtitle = "Median income $52K in year 2013",
       fill = "Income Levels")+
  ylab("Income Level (proportions)")+
  xlab("General Health (proportions)")+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())+
  scale_fill_manual(values = c("below median" = "red", "above median" = "blue"))
```
\

Conclusion: \

- the health condition is clearly correlated with income level\
- lower income shows less good health conditions\
- higher income shows better than good health conditions\

However, since this sample is not from a controlled experiment, we cannot draw the conclusion that a less good health is caused by lower income.\
\



#### Question 1a: Probability of good health given a high income?

Review contingency table with margins (sums):

```{r contigency table}
ct <- tb_cont_table2_agg_q1
ct
```

Question can be formulated as `P(good health | high income)`:

```{r probability good health given high income}
# P(good health | high income) = P(A|B)
p_a_given_b <- ct[2,1]/ct[2,3]
p_a_given_b
```

If someone has a higher income above average the probability of good health is 0.66 (or 66%).\
\


#### Question 1b: Probability of high income given a good health?

Question can be formulated as `P(high income | good health)`\
Can be calculated with Bayes Theorem: `P(B|A) = (P(A|B) * P(B)) / P(A)`\
\

Calculate from contingency table: P(B)   = P(high income)

```{r calculate probability of high income}
# P(high income) = P(B)
p_b <- ct[2,3]/ct[3,3]
p_b
```

Calculate from contingency table: P(A)   = P(good health)

```{r caclulate probability of good health}
# P(good health) = P(A)
p_a <- ct[3,1]/ct[3,3]
p_a
```

Calculate by Bayes Theorem: P(B|A) = P(A|B) * P(B) / P(A)

```{r calculate cond prob of high income given good health}
# P(high income | good health) = P(B|A)
p_b_given_a <- p_a_given_b * p_b / p_a
p_b_given_a
```
If someone is in category "more than good health" the probability to be in the higher income group is 0.56 (or 56%).
\
\

**Question 1c: Test for independence between income level and general health**

To proof that event A (health) and B (income) are independent from each other the following formulas must hold true:

P(A|B) = P(A):     P(good health | high income) = P(good health), and\
P(B|A) = P(B):     P(high income | good health) = P(high income) 

```{r test if event A and B are independent}
# Test that events A and B are independent
p_a_given_b == p_a
p_b_given_a == p_b
```
Result:\

- P(A|B) <> P(A) and P(B|A) <> P(B) are booth FALSE\
- Therefore, health condition and income level are not independent
\
\

**Conclusion of question 1:**

In this question we utilized two categorical variables: income and general health. From contingency tables we found that income and general health are correlated. We visualized the relationship by mosaic plots in detail and in aggregated form.\

From an aggregated contingency table we calculated conditional probabilities:\

- Probability of good health given high income is 66%\
- Probability of high income given good health is 56%\

We further verified that both variables are dependent from each other\
\


**Cleaning up Global Environment**

```{r}
rm(ct,
   df_cont_table1_agg_q1,
   df_cont_table1_q1,
   df_q1,
   df_select_q1,
   tb_cont_table2_agg_q1,
   tb_cont_table1_agg_q1,
   tb_cont_table1_q1,
   p_a, p_a_given_b, p_b, p_b_given_a)
```

\
\
\


### Research question 2:

Is general health (`genhth`) correlated with physical activity (`pa1min_`) and how does it change by gender (`sex`)?\

a. Describe physical activity statistically\
b. Find the relationship between general health and physical activity\
c. Is there a difference in this relationship (b) in respect to gender?\

Here we are going to compare numerical values (`pa1min_`) across categories (`genhth`) and (`sex`).\ First, we will investigate the numerical variable (`pa1min_`) statistically, by means of histogram, box-plot, and summary statistics.\
Second, we will compare the numerical values across category (`genhth`) by side-by-side box-plots.\
Finally, we will compare the numerical values across categories (`genhth`) and (`sex`) by 2-dimensional side-by-side box-plots\
\


#### Preparing the data set and data cleaning

**Reduce the number of variables**: Extract variables `genhth`, `sex` and `pa1min_` 

```{r data cleaning q2}
df_select_q2 <- df_raw %>% 
  select(genhlth, sex, pa1min_)
```
\

**Remove NAs**

```{r remove NAs variables q2}
# remove NAs in variables
df_q2 <- df_select_q2 %>% 
  group_by(genhlth,sex) %>% 
  filter(!is.na(genhlth), !is.na(sex), !is.na(pa1min_)) %>% 
  arrange(desc(pa1min_))
```
\

#### Exploring the data set
\

**Inspecting the data** for variable `pa1min_`:

Calculate the statistical summary for variable physical activity `pa1min_` in count of times per month:

```{r stat summary raw data q2}
# get stat summary
df_stat_q2 <- df_q2 %>%
  group_by(.) %>% 
  summarise(
    mean_pa = round(mean(pa1min_),1),
    sd_pa = round(sd(pa1min_),1),
    median_pa = round(median(pa1min_),1),
    iqr_pa = round(IQR(pa1min_),1),
    q1_pa = round(quantile(pa1min_, 0.25),1),
    q3_pa = round(quantile(pa1min_, 0.75),1),
    min_pa = round(min(pa1min_),1),
    max_pa = round(max(pa1min_),1),
    count_pa = n()
  )

# Pivot data frame for print out
df_stat_q2_t <- df_stat_q2 %>% 
  pivot_longer(c(`mean_pa`,
                 `sd_pa`,
                 `median_pa`,
                 `iqr_pa`,
                 `q1_pa`,
                 `q3_pa`,
                 `min_pa`,
                 `max_pa`,
                 `count_pa`),
               names_to = "stat values", values_to = "values")
df_stat_q2_t
```
\

**Visualize raw data distribution by histogram and box-plot**

Histogram:

```{r histogram q2}
df_q2 %>% 
  ggplot()+
  geom_histogram(aes(x=pa1min_), binwidth = 60)
```
\

Highly right-skewed distribution indicates the existence of plenty of extreme values on the right\
\

Box-Plot:

```{r box plot q2}
df_q2 %>% 
  ggplot(aes(x="", y=pa1min_))+
  geom_boxplot()+
  xlab("")
```
\

The max whisker can be calculated by "Q3 + 1.5 IQR". Values more extreme than the max. whisker may be outliers and be removed

```{r upper whisker q2}
# calculate max whisker
max_whisker_pa <- as.numeric(df_stat_q2_t[6,2] + 1.5 * df_stat_q2_t[4,2])
max_whisker_pa
```

The box-plot shows plenty of values above the max whisker (1072). There are also values above the possible minutes per week (i.e. 10,080min). The max whisker of 1072min per week translates into 2h33m per day. We make the assumption that the maximal physical activity of up to 4 hours per day (i.e. 1680 min per week) is plausible. Therefore, we will set the threshold for "cut-off" to 1680.\
\


**Remove outliers**

Cutting off values above the assumed max value 1680 minutes:

```{r remove outliers q2}
cut_off_value_pa <- 4*60*7

# remove outliers
df_q2_cleaned <- df_q2 %>% 
  filter(pa1min_ <= cut_off_value_pa)
```
\
\


**Inspecting the data** for categorical variables by bar-plots\

General health:

```{r distribution cat values genhlth q2b}
# Bar chart of category genhelth (COUNT)
df_q2_cleaned %>% 
  ggplot(aes(x = genhlth)) +
  geom_bar(fill="gray", color="white")+
  labs(title = "General health: Distribution of counts")
```
\
Highest counts are in categories "very good" and "good"\
\

Sex:

```{r distribution cat values income levels q2}
# Bar chart of category income levels (COUNT)
df_q2_cleaned %>% 
  ggplot(aes(x = sex)) +
  geom_bar(aes(x = sex, y = (after_stat(count))/sum(after_stat(count))),fill="gray", color="white")+
  labs(title = "Sex: Proportions")+
  xlab("sex")+
  ylab("proportion")
```
\
The data set contains about 42% male and 58% female\
\
\

#### Question 2a: Describe physical activity statistically\

Calculate a **statistical summary** for variable physical activity `pa1min_` from the cleaned data set

```{r stat summary raw data q2 cleaned}
# get stat summary
df_stat_q2_cleaned <- df_q2_cleaned %>%
  group_by(.) %>% 
  summarise(
    mean_pa = round(mean(pa1min_),1),
    sd_pa = round(sd(pa1min_),1),
    median_pa = round(median(pa1min_),1),
    iqr_pa = round(IQR(pa1min_),1),
    q1_pa = round(quantile(pa1min_, 0.25),1),
    q3_pa = round(quantile(pa1min_, 0.75),1),
    min_pa = round(min(pa1min_),1),
    max_pa = round(max(pa1min_),1),
    count_pa = n()
  )

# Pivot data frame for better visualization
df_stat_q2_t_cleaned <- df_stat_q2_cleaned %>% 
  pivot_longer(c(`mean_pa`,
                 `sd_pa`,
                 `median_pa`,
                 `iqr_pa`,
                 `q1_pa`,
                 `q3_pa`,
                 `min_pa`,
                 `max_pa`,
                 `count_pa`),
               names_to = "stat values", values_to = "values")
df_stat_q2_t_cleaned
```

Histogram (cleaned data):

```{r histogram cleaned q2}
df_q2_cleaned %>% 
  ggplot()+
  geom_histogram(aes(x=pa1min_), fill="darkgray", color ="lightblue", binwidth = 60)+
  labs(title = "Distribution: Physical activities in min per week",
       subtitle = "Bin width = 60 min",
       fill = "Income Levels")+
  ylab("Counts")+
  xlab("Physical Activity [min/week]")

```
\
\

Box-Plot (cleaned data):

```{r box plot cleaned q2}
df_q2_cleaned %>% 
  ggplot(aes(x="", y=pa1min_))+
  geom_boxplot()+
    labs(title = "Distribution: Physical activities in min per week")+
  ylab("Physical Activity [min/week]")+
  xlab("")
```
\
Summary:\

- 50% of values range from 105min to  450min or IQR = 345min\
- The data are right-skewed with median = 233 and IQR = 345\

This cleaned data set is plausible and will be used to answer the next questions.

\
\


#### Question 2b: How are physical activity and general health related

Comparing numerical variable `pa1min-` by categorical variable `genhlth` using box-plots

```{r box plot q2b, fig.width=8, fig.asp=0.6}
# box plot for physical activity (numerical) by general health (factor)
df_q2_cleaned %>% 
  ggplot()+
  geom_boxplot(aes(genhlth, pa1min_))+
  scale_x_discrete(limits = rev(levels(factor(df_q2_cleaned$genhlth))))+
  coord_flip()+
  labs(title = "Distribution: Physical activities by general health")+
  ylab("Physical Activity [min/week]")+
  xlab("General Health")
```
\

Conclusion:\

- The graph shows a trend that the health status is related to physical activity\
- The median (dark vertical line in the box) is moving from left to right if we move upwards from "Poor" to "Excellent" health condition. The more physical activity the better the health condition.\
\
However, we can not conclude from the data that more physical activity causes an improvement in health.\
\
\

#### Question 2c. Is there a difference in this relationship (b) in respect to gender

Finally, we like to investigate if there is a significant difference between male and female in respect of physical activity and general health. We will introduce a third variable `sex` and visualize this by a 2-dimensional box-plot


```{r box plot q2c, fig.width=8, fig.asp=0.6}
# box plot for physical activity (numerical) by general health (factor) and sex (factor)

df_q2_cleaned %>% 
  ggplot()+
  geom_boxplot(aes(genhlth, pa1min_))+
  scale_x_discrete(limits = rev(levels(factor(df_q2_cleaned$genhlth))))+
  coord_flip()+
  facet_grid(.~sex)+
  labs(title = "Distribution: Physical activities by general health and sex")+
  ylab("Physical Activity [min/week]")+
  xlab("General Health")
```
\

Conclusion:\

- there is an almost identical trend that more physical activity is correlated to better health condition in group male and female\
- male show slightly more physical activity than female\
\


Conclusion of question 2:

In this questions we utilized three variables: one numerical variable for physical activity in minutes per week, and two categorical variables for general health and sex. The numerical data had some implausible data (outliers) and needed to be filtered out. The comparison of all variable was done with box-plots which allow an easy comparison over multiple categories.

As expected, general health condition is correlated with more physical activities. We couldn't find a difference in this relation between male and female. The conclusion is only valid for correlation but not for causation.



**Cleaning up Global Environment**

```{r}
rm(df_q2,
   df_q2_cleaned,
   df_select_q2,
   df_stat_q2,
   df_stat_q2_cleaned,
   df_stat_q2_t,
   df_stat_q2_t_cleaned
  )
```
\
\

### Research question 3:


Is physical activity (`pa1min_`) correlated with **BMI (Body Mass Index)** (`X_bmi5`) and how does it change with general health (`genhlth`) and gender (`sex`)?\

a. Describe BMI statistically\
b. Find the relationship between physical activity and BMI\
c. Is there a difference in this relationship (b) in respect to general health and gender?\


Here we are going to compare two numerical variables (`pa1min_`) and (`X_bmi5`) across the categorical variables (`genhth`) and (`sex`). First, we will investigate the numerical variable (`X_bmi5`) statistically, by means of descriptive summary, histogram and box-plot. Second, we will calculate and visualize the correlation between the two numerical variables (`pa1min_` and `X_bmi5`) by a scatter plot. Finally, we will for differences by categories (`genhth`) and (`sex`) using multiple scatter plots\
\


#### Preparing the data set and data cleaning

**Reduce the number of variables**: Extract variables `sex`, `genhlth`, `pa1min_`, `X_bmi5` 

```{r data cleaning q3}
df_select_q3 <- df_raw %>% 
  select(sex, genhlth, pa1min_, X_bmi5)
```
\

**Remove NAs**

```{r remove NAs variables q3}
# remove NAs in variables
df_q3 <- df_select_q3 %>% 
  group_by(sex, genhlth) %>% 
  filter(!is.na(sex), !is.na(genhlth), !is.na(pa1min_), !is.na(X_bmi5)) 
```
\
\


#### Question 3a: Describe variable BMI statistically

The BMI values in variable `X_bmi5` includes 2 decimals. First, we convert the values to BMI convention by division with 100

```{r convert BMI}
df_q3 <- df_q3 %>% 
  mutate(bmi = X_bmi5/100)
```
\


Summary statistics:

```{r stat summary raw data q3 - BMI }
# get stat summary
df_stat_q3_bmi <- df_q3 %>%
  group_by(.) %>% 
  summarise(
    mean_bmi = round(mean(bmi),1),
    sd_bmi = round(sd(bmi),1),
    median_bmi = round(median(bmi),1),
    iqr_bmi = round(IQR(bmi),1),
    q1_bmi = round(quantile(bmi, 0.25),1),
    q3_bmi = round(quantile(bmi, 0.75),1),
    min_bmi = round(min(bmi),1),
    max_bmi = round(max(bmi),1),
    count_bmi = n()
  )

# Pivot data frame for better print out
df_stat_q3_bmi_t <- df_stat_q3_bmi %>% 
  pivot_longer(c(`mean_bmi`,
                 `sd_bmi`,
                 `median_bmi`,
                 `iqr_bmi`,
                 `q1_bmi`,
                 `q3_bmi`,
                 `min_bmi`,
                 `max_bmi`,
                 `count_bmi`),
               names_to = "stat values", values_to = "values")
df_stat_q3_bmi_t
```
\

Histogram:

```{r histogram q3 - BMI raw}
df_q3 %>% 
  ggplot()+
  geom_histogram(aes(x=bmi), binwidth = 2)+
  labs(title = "Distribution of BMI")
```
\

Slightly right-skewed with some extreme values on the right of the x-axis (mean = 27.4, sd = 5.7, max = 95.1)\
\

Box-Plot:

```{r box plot q3 - BMI raw}
df_q3 %>% 
  ggplot(aes(x="", y=bmi))+
  geom_boxplot()+
  labs(title = "Distribution of BMI")+
  xlab("")
```
\

Calculating the max whisker:

The max whisker is calculated by "Q3 + 1.5 IQR". Values more extreme than the max. whisker maybe outliers and can be deleted

```{r max whisker q3 - BMI raw}
# calculate max whisker
max_whisker_bmi <- as.numeric(df_stat_q3_bmi_t[6,2] + 1.5 * df_stat_q3_bmi_t[4,2])
max_whisker_bmi
```

The box-plot shows extreme values above the max whisker (40). We make the assumption that the maximal BMI is 50. Every value above the threshold (50) will be cut off\
\


**Remove outliers** from `bmi` and  from `pa1min_`

We will remove from BMI values > 50 and from physical activity values > 1680 (see question 2a):

```{r remove outliers q3}
cut_off_value_bmi <- 50

# remove outliers
df_q3_cleaned <- df_q3 %>% 
  filter(bmi <= cut_off_value_bmi, pa1min_ <= cut_off_value_pa) 
```
\
\


#### Question 3b: Find the relationship between physical activity and BMI

To visualize two numerical variables we will use a scatter plot with physical activity as input variable (x-axis) and BMI as response variable (y-axis). Due to many data points (about 300K) the scatter plot will become too crowded. We, therefore, will reduce the number of data points to 15K (i.e. 5%) by random sampling function `sample()`.

```{r reduce number of data points}

# reduce number of data points (observation) by random sampling
df_q3_sampled <- df_q3_cleaned[sample(nrow(df_q3_cleaned), nrow(df_q3_cleaned)*0.05), ]

```
\

Scatter plot:

```{r scatter plot q3b original, eval=FALSE, fig.asp=0.6, fig.width=8, warning=FALSE, include=FALSE}
# create scatter plot original 
# use log10 for pa1min_

df_q3_cleaned %>% 
  ggplot(aes(x = pa1min_, y = bmi))+
  geom_point(alpha=0.1, color="blue") +
  scale_x_log10()+
  labs(title = "Relationship between Physical Activity and BMI",
       subtitle = "Original data, n = 296,506")+
  xlab("Physical Activity in log10 [min/week]")+
  ylab("BMI")
```


```{r scatter plot q3b, fig.width=8, fig.asp=0.6, warning=FALSE}
# create scatter plot - sampled
# use log10 for pa1min_

df_q3_sampled %>% 
  ggplot(aes(x = pa1min_, y = bmi))+
  geom_point(alpha=0.1, color="blue") +
  scale_x_log10()+
  labs(title = "Relationship between Physical Activity and BMI",
       subtitle = "Data sampled by 5% from original, n_sampled = 14,825")+
  xlab("Physical Activity in log10 [min/week]")+
  ylab("BMI")
```
\

Calculating the correlation coefficient between both variables `pa1min_` and `bmi`

```{r cor coeff q3 - bmi}
# calculate the correlation coefficient between pa1min_ and bmi
cor_pa_bmi <- cor(df_q3_sampled$pa1min_, df_q3_sampled$bmi)
cor_pa_bmi
```

Conclusion:\

- The scatter plot does not show any correlation\
- The correlation coefficient is -0.08.\
\
\

#### Question 3c: Is there a difference in this relationship (b) in respect to general health and sex?\

The sample data will be grouped by general health condition and sex, using the `facet_grid()` function.

```{r q3c, fig.width=8, fig.asp=0.6, warning=FALSE}
# Relationship between physical activity and BMI grouped by health condition and sex
# fig.width=12, fig.asp=0.6
df_q3_sampled %>% 
  ggplot(aes(x = pa1min_, y = bmi))+
  geom_point(alpha=0.1, color="blue") +
  facet_grid(sex~genhlth)+
  scale_x_log10()+
  labs(title = "Relationship between Physical Activity and BMI",
       subtitle = "Sub Grouping: General Health Status (horizontal), Gender (vertical)")+
  xlab("Physical Activity in log10 [min/week]")+
  ylab("BMI")
```
\
Conclusion:\

- Non of the group shows a correlation between Physical Activity and BMI\
- The better the health condition the smaller the spread around the center (BMI average and Physical Activity average)\
- The female groups show a slightly larger spread in BMI around the average than male\
\
\

**Conclusion of question 3:**

In this questions we utilized four variables, which are two numerical variables "physical activity" and "BMI" and two categorical variables "general health" and "sex". For the comparison of two numerical variable a scatter plot was used. For better data visualization the data amount had to be reduced using a random sampling function from R. Furthermore, the x scale "physical activity" had to be transformed by LOG10 function.

Unfortunately, the data set did not reveal any correlation between these two variables, as it could have been expected. We are not sure if we interpreted the numerical values for BMI correctly. Further information from the official site is required. 

The relationship between the numerical variables "physical activity" and "BMI, was further investigated for categories "general health" and "sex". Non of the group showed any correlations between the "physical activity" and "BMI".
However, in the group of "Excellent Health Condition" the "physical activity" and "BMI" data are less spread around the center value.\
\

**References**

- [Data Analysis with R Specialization (Coursera)](https://www.coursera.org/specializations/statistics)






